---
title: "Lab for GEOG 3023: Statistics for Geography"
subtitle: "Lab 4: Statistical Relationships in R" 
author: Elliot Whitehead
output: 
  html_document: default
---

```{r setup, include=FALSE}

# Setup the environment
library(knitr)
knitr::opts_chunk$set(fig.align='center',fig.width=10, fig.height=6, fig.path='Figs/',  warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE)

r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

```



Complete all **5 Questions**, and submit the finished Rmd and knitted HTML file onto Canvas. Don't forget to change name field in 
the beginning to your first and last name. 

This lab is based on a lab developed by Dr. Matthew Crump at CUNY Brooklyn (https://crumplab.github.io/statisticsLab/)


# Introduction


In lecture, we have been discussing the idea of correlation. This is the idea that two things that we measure can be 
somehow related to one another. For example, your personal happiness, which we could try to measure say with a questionnaire,
might be related to other things in your life that we could also measure, such as number of close friends, yearly salary, 
how much chocolate you have in your bedroom, or how many times you have said the word Nintendo in your life. Some of the 
relationships that we can measure are meaningful, and might reflect a causal relationship where one thing causes a change 
in another thing. Some of the relationships are spurious, and do not reflect a causal relationship.

In this lab you will learn how to compute correlations between two variables in 'R', and then ask some questions about the correlations that you observe.

## Learning Goals

  1.  Compute Pearson’s r between two variables with R
  2.  Simple linear regression with R
  3.  Discuss the possible meaning of correlations that you observe


## Getting started

### Load packages

In this lab we will continue use the 'dplyr' package used for the last week. 

Let's load the packages:

```{r load-packages, message=FALSE}
library(dplyr)
library(ggplot2)
```

### Data

We use data from the [World Happiness Report](https://worldhappiness.report/). The World Happiness Report is a 
landmark survey of the state of global happiness that ranks 166 countries by how happy their citizens perceive themselves to be.  
A csv of the data for the year 2021 is included in the lab materials. The data includes annual survey results 
from the year 2008 to 2020 for each of the 166 countries. Beware that there is missing values ('NA') for some of the columns.

Let's load the data:

```{r, echo=T, results=F}

# Make sure the data are in the right folder
# 
whr_data <- read.csv('./WHR2021.csv')


# Now browse the data and see what columns the data have. The names of the column should explain themselves
names(whr_data)

# Show the first few rows of the data frame.
head(whr_data)

```

You should be able to see that there is data for many different countries, across a few different years. 
There are lots of different kinds of measures, and each are given a name that should explain themselves. 
You can find out more info here: https://worldhappiness.report/faq/


# Getting Familiar with the Data

## Life Satisfaction over time

One of the most interesting columns is the ``Life.Ladder'' variable. This is a descriptor of life satisfaction and happiness. 
From the WHR website: ``This is called the Cantril ladder: it asks respondents to think of a ladder, with the best possible 
life for them being a 10, and the worst possible life being a 0. They are then asked to rate their own current lives on that 0 to 10 scale.'' 
Then WHR calculates the mean rating for each country and that's the value you see in ``Life.Ladder''

To start, let's explore the changes of the Life.Ladder over the years for the United States.

```{r}

# By using dplyr, we can easily select the rows and columns in the dataframe that we want
# Then save it to a new dataframe called "whr_data_us"
whr_data_us <- whr_data %>%
  select(country, year, Life.Ladder) %>% #select the interested columns
  filter(country == 'United States')  # only choose United States

# Plot
ggplot(whr_data_us) + aes(x=year, y=Life.Ladder) + geom_line() +
  ggtitle("US Mean Life.Ladder, 2008 - 2020")

```

## Life Satisfaction over time: comparing countries

It looks like life satisfaction has dropped a bit over the last decade in the United States, but seen a small rebound
in the last few years. Let's look at another country to compare between the two. In this case, we'll use Germany

```{r}
# By using dplyr, we can easily select the rows and columns in the dataframe that we want
# Then we save it to a new dataframe called "whr_data_us_germany"
whr_data_us_germany <- whr_data %>%
  select(country, year, Life.Ladder) %>% #select the interested columns
  filter(country == 'United States' | country == 'Germany')  # choose US AND Germany

# Plot, using "country" as the color aesthetic
ggplot(whr_data_us_germany) + aes(x=year, y=Life.Ladder, color=country) + geom_line() +
  ggtitle("Mean Life.Ladder, 2008 - 2020")


```


### Your turn: 

For question 1, you'll look at the ``Freedom.to.make.life.choices'' variable but for
two DIFFERENT countries (NOT the ones above)

**Q1: (15 pts)** Similar to the analysis above, choose two new countries you're interested
in and make a line plot comparing their ``Freedom.to.make.life.choices'' variable over time

Hint: To see what countries are options, click on the ``whr_data'' object in the Environment pane (top right)
Notice that not all countries have all years, so pick two countries that have data for the whole time period

```{r}
# First, use select() and filter() to create a new dataframe that has only the
# countries and columns you're interested in
whr_data_life_freedom <- whr_data %>%
  select(country, year, Life.Ladder) %>%
  filter(country == 'Turkey' | country == 'Denmark')

# Then, plot!
ggplot(whr_data_life_freedom) + aes(x=year, y=Life.Ladder, color=country) + geom_line() + ggtitle("Average Life Ladder, by Year")


```

# Calculating Correlation

Does a countries measure for “Freedom.to.make.life.choices” correlate with that countries measure for "Life.Ladder"?

Let’s find out. We calculate the correlation, and then we make the scatter plot.

```{r}

cor(whr_data$Freedom.to.make.life.choices, whr_data$Life.Ladder)
ggplot(whr_data) + aes(x=Freedom.to.make.life.choices, y=Life.Ladder) + geom_point()

```

What happened here? We can see some dots, but the correlation was NA (meaning missing values or undefined). 
This occurred because there are some missing data points in the data.
We can remove all the rows with missing data first, then do the correlation. 

We will do this a couple steps, first creating our own dataframe with only the 
numbers we want to analyze. Similar with the previous case, we can select the
columns we want to keep using `select()`. Then we use `filter()` to remove the
rows with NAs (! means NOT, so !is.na() means NOT NA)

```{r}

# Remove NA rows and save it as a new dataframe
whr_clean <- whr_data %>%
               filter(!is.na(Freedom.to.make.life.choices), # ! means NOT, so !is.na() means NOT NA
                      !is.na(Life.Ladder),
                      !is.na(Social.support))

whr_clean_Freedom_LifeLadder <- whr_clean %>%
  select(country, Freedom.to.make.life.choices, Life.Ladder)

# Calculate correlation!
cor(whr_clean_Freedom_LifeLadder$Freedom.to.make.life.choices,
    whr_clean_Freedom_LifeLadder$Life.Ladder)

```

Now we see the correlation is .528. That's a pretty strong positive correlation!

## Adding line of best fit
Although the scatter plot shows the dots are everywhere, it generally shows
that as Freedom to make life choices increases in a country, that country's
life satisfaction (Life.Ladder) also increase. This is a positive correlation. 
Let’s do this again and add the best fit line, so the trend is more clear. 
The function to fit the line or simple regression is `lm()`, and the way to 
use it is `lm(y~x)` where `y~x` is the `formula` to be fitted, `y` is the 
response or dependent variable, `x` is the explanatory/independent variable, 
and `~` sign is a separator. Type `?lm` to see more about how to use this function.

```{r}

# first make the scatter plot
ggplot(whr_clean_Freedom_LifeLadder) + aes(x=Freedom.to.make.life.choices, y=Life.Ladder) + geom_point() +
  geom_smooth(method='lm', formula=y~x) # Then add the line of best fit! It's as easy as that!

```


# Linear Modeling

In the last code block, we had ggplot fit a linear model and add that line to
our plot (with some error margins)

Here, we'll fit the same linear model and investigate it more

To fit a linear model in R, you use the `lm()` command. Inside the command,
you give it a `formula` with the variables you want to model: `lm(y~x)`
Notice how the y variable comes FIRST and the x comes SECOND. 
How you can read this is `target variable ~ predictor variable`

Let's fit a linear model predicting Life.Ladder (target variable) using
Freedom.to.make.life.choices (predictor variable)
```{r}
linear_model <- lm(whr_data$Life.Ladder~whr_data$Freedom.to.make.life.choices)

```


What `lm(y~x)` does is to fit the simple linear regression model `y = a + bx`.
In other words, `lm()` can estimate the coefficients `a` (intercept) and `b` 
(slope). We can assess the intercept and slope like the following:

```{r}

# show the summary of the fitted linear model
summary(linear_model)


# We can access the fitted intercepts and slope values
linear_model$coefficients # this is a vector with two entries, the first one is intercept and other is slope


linear_model$coefficients[1] # is the intercept
linear_model$coefficients[2] # is the slope

```

# Your turn: Questions 2-5 ask you to investigate the relationship between two variables.

**Q2: (20 pts)** One of the other columns in the dataset `whr_data`, is `Social.support`. Please calculate the correlation coefficient between Social.support and Life.Ladder.
Be careful of the `NA` values in the columns as shown earlier. 

Then, interpret what the correlation coefficient means answering these questions:

1. What is the direction of the correlation?

2. How strong is it?

3. Is Social.support more or less correlated to Life.ladder than Freedom.to.make.life.choices is?

**Your answer:**

```{r}

# select data
whr_clean_Social_Support <- whr_clean %>%
  select(country, Social.support, Life.Ladder)

# assess correlation
cor(whr_clean_Social_Support$Social.support,
    whr_clean_Social_Support$Life.Ladder)
```
#### "Social support" has a stronger positive correlation (.71) to "Life Ladder" than "Freedom to make life choices" does (.53).

Discussion (2-3 sentences): 


**Q3: (15 pts)**  Following the previous question Q2, please draw a scatter plot between columns `Life.ladder` 
and `Social.support` (`Social.support` as x axis and `Life.ladder` as y axis) and add the regression line.
**Your answer:**

```{r}
# first make the scatter plot
ggplot(whr_clean) + aes(x=Social.support, y=Life.Ladder) + geom_point()

```


**Q4: (20 pts)** Following the previous question Q3, fit a simple linear regression model to predict 
`Life.Ladder` using `Social.support` as our predictor. In other words, `Life.ladder` is `y` and `Social.support` is `x`. 
The regression line can be described as `y = a + bx`. After fitting the model, answer the following questions:

1. What are fitted values for `a` and `b`? 

2. Is the value of `b` a positive or negative and what does it mean?

3. Look back to the linear modeling results for `Freedom.to.make.life.choices`. Based on the values of `b` for 
the two linear models, which variable has a larger impact on Life.Ladder? In other words, per unit increase in
Freedom.to.make.life.choices and Social.support, which results in a larger increase in Life.Ladder? 


**Your answer:**

```{r}
# create linear model
social_support_linear_model <- lm(whr_clean_Social_Support$Life.Ladder~whr_clean_Social_Support$Social.support)

summary(social_support_linear_model)

# display the linear model coefficients
sprintf("Intercept: %f", social_support_linear_model$coefficients[1])
sprintf("Slope: %f", social_support_linear_model$coefficients[2])

# plot that shit
ggplot(whr_clean) + aes(x=Social.support, y=Life.Ladder) + geom_point() +
  geom_smooth(method='lm', formula=y~x)
```

### Discussion (3-4 sentences): 

#### The fitted value for `a` is `0.047697` and the fitted value for `b` is `6.668080`. The slope is positive, and quite steep, meaning that the correlation between Social.support and Life.Ladder data is quite strong, meaning that Social.support has a greater impact on Life.Ladder than Freedom.to.make.life.choices.


**Q5: (30 pts)** This question is more open ended and proceeds in multiple steps. 

So far, we've been focusing on the `Life.Ladder` variable. But there are other
other measures of happiness in the dataset, especially `Positive.affect` and 
`Negative.affect`. Positive affect is the frequency of positive feelings
that participants report. Negative affect is the frequency of negative feelings
they report. You can google "Positive affect WHR" and "Negative affect WHR"
to find out more.

Here's what you'll do:
*A* Create a new variable called `Total.affect` using dplyr's mutate() function
(you'll probably need to Google to figure out how!)
*B* Choose one other variable to analyze in relation to your new variable (can
be anything except for Positive and Negative affect)
*C* Create a scatter plot with `Total.affect` on the y-axis and your other 
variable on the x axis. Add the line of best fit
*D* Calculate Pearson's correlation coefficient in R
*E* Fit a linear model (remember, put Total.affect first since it's your 
y-variable, e.g. lm(whr_data$Total.affect~Other.variable))
*F* In 3-4 sentences, summarize your findings! Is there a relationship between
your variable and Total.affect? Is it positive or negative? Is it strong or weak?
What do you think this tells us about happiness? Do you think this is a causal
relationship or merely correlative? 


```{r}

# Put your code here


# Part A


# Part C


# Part D


# Part E


```

Discussion (3-4 sentences): 

Done! Make sure to submit both your .Rmd and knitted .html file to Canvas!
