---
title: "GEOG 3023: Statistics for Geography"
subtitle: "Lab 7: T Tests"
author: Elliot Whitehead
output: 
  html_document:
    css: "lab.css"
---

```{r setup, include=FALSE}

# Setup the environment
library(knitr)
knitr::opts_chunk$set(fig.align='center',fig.width=10, fig.height=6, fig.path='Figs/',  warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE)

r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

```

<div class="instructions">

Please complete the following questions and submit the finished Rmd and
HTML file onto Canvas. 

Don't forget to change name field in the beginning to your **first and last name**. 

This lab is modified based on [OpenIntro](https://www.openintro.org/stat/labs.php) lab.

</div>

## Getting Started

### IMPORTANT NOTE

**For this entire lab, we'll use a significance level (alpha) of 0.05**

### Load packages

In this lab we will explore the data using the `dplyr` package and
visualize it using the `R` basic graphics for data visualization. The data
 ('nc.csv') is enclosed. 

Let's first load the package.

```{r load-packages, message=FALSE}
library(dplyr)
library(ggplot2)
```

### The data

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

Load the `nc` data set into our workspace.

```{r load-data}
nc<-read.csv('./nc.csv')
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | ---------------------------------------------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

As a first step in the analysis, we should take a look at the variables in
the dataset. This can be done using the following commonly used commands:

```{r str}

# show the structure of the data frame
str(nc)

# display the first few lings of the data frame
head(nc)

# return the dimension of the data frame
dim(nc)

# return the column names of the data frame
colnames(nc)
```

As you review the variable summaries, consider which variables are categorical and which are numerical. For numerical variables, are there outliers? If you aren't sure or want to take a closer look at the data, make a graph.

## Exploratory data analysis

We will first start with analyzing the weight gained by mothers throughout the 
pregnancy: `gained`.

Using visualization and summary statistics, describe the distribution of weight 
gained by mothers during pregnancy. The `summary` function can also be
useful. Note there are missing values in the dataset. 

```{r summary}
summary(nc$gained)
summary(nc$habit)
```

Next, consider the possible relationship between a mother's smoking habit and the weight of her baby. Plotting the data is a useful first step because it helps us quickly visualize trends, identify strong associations, and develop research questions.

<div class="question">
**Q1:** Make side-by-side boxplots of `habit` and `weight`. Which of the following is false about the relationship between habit and weight?

* Median birth weight of babies born to non-smoker mothers is slightly higher than that of babies born to smoker mothers. 
* Range of birth weights of babies born to non-smoker mothers is smaller than that of babies born to smoker mothers. 
* The IQRs of the distributions are roughly equal. 

**Your answer:**


```{r habit-weight-box}
# Clean data from NA values for habit
nc_clean <- na.omit(nc)

ggplot(nc_clean) + aes(x=habit, y=weight, fill=habit) + geom_boxplot()
```

<br>
**False: Range of birth weights of babies born to non-smoker mothers is smaller than that of babies born to smoker mothers.**

</div>

The box plots show how the medians of the two distributions compare, but we can also compare the means of the distributions using the following to 
first group the data by the `habit` variable, and then calculate the mean
`weight` in these groups using the `mean` function.

```{r by-means}
nc %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis 
test.

## Hypothesis test of population mean

### One-Sample T-Test

<div class="question">
**Q2**: The mean baby weight in the US is 7.5 lbs. We want to test if, based
on this sample, the North Carolina mean weight is different from the US average (two-tailed).
What are the appropriate null and alternative hypotheses for this test?

* $H_0: \mu_{NC} = 7.5$; $H_A: \mu_{NC} > 7.5$ 
* $H_0: \mu_{NC} = 7.5$; $H_A: \mu_{NC} \ne 7.5$ 
* $H_0: \bar{x}_{NC} = 7.5$; $H_A: \bar{x}_{NC} >7.5$ 
* $H_0: \bar{x}_{NC} = 7.5$; $H_A: \bar{x}_{NC} < 7.5$ 
* $H_0: \mu_{NC} \ne 7.5$;  $H_A: \mu_{NC} = 7.5$ 


**Your Answer:**

$H_0: \mu_{NC} = 7.5$; $H_A: \mu_{NC} \ne 7.5$

</div>

<div class="question">

**Q3**: Run a 1-sample t-test to determine if, based on this sample, the mean 
North Carolina baby weight is different from the US average. Interpret the results
at a 0.05 significance level. Is the NC baby weight different from the US avg?

**Your answer:**


```{r}
t.test(nc_clean$weight, mu=7.5)
```
**Our p-value is much smaller than our alpha of 0.05, so we can reject the null
hypothesis and assume that the mean NC baby weight is different than from the US avg.**
</div>


### Two-Sample T-Test

<div class="question">
**Q4:**  What are the hypotheses for testing if the average weights of babies born to smoking and non-smoking mothers are different?

* $H_0: \mu_{smoking} = \mu_{non-smoking}$; $H_A: \mu_{smoking} > \mu_{non-smoking}$ 
* $H_0: \mu_{smoking} = \mu_{non-smoking}$; $H_A: \mu_{smoking} \ne \mu_{non-smoking}$ 
* $H_0: \bar{x}_{smoking} = \bar{x}_{non-smoking}$; $H_A: \bar{x}_{smoking} > \bar{x}_{non-smoking}$ 
* $H_0: \bar{x}_{smoking} = \bar{x}_{non-smoking}$; $H_A: \bar{x}_{smoking} > \bar{x}_{non-smoking}$ 
* $H_0: \mu_{smoking} \ne \mu_{non-smoking}$;  $H_A: \mu_{smoking} = \mu_{non-smoking}$ 

**Your answer:**

$H_0: \mu_{smoking} = \mu_{non-smoking}$; $H_A: \mu_{smoking} \ne \mu_{non-smoking}$
</div>

Before running our test, we first need to know the weights for the `smoker` and `nonsmoker`:

```{r}
smoker <- nc %>%
    filter(habit=='smoker')

nonsmoker <- nc %>%
    filter(habit=='nonsmoker')

```

<div class="question">

**Q5:** Run a two-sample t-test comparing baby weights of smoking and non-smoking mothers.
Based on the output of, what is the conclusion of the hypothesis test? 
Are the weights of babies for smoking mothers less than, the same as, or greater than 
for non-smoking mothers? (After your test, you can calculate the mean of each sample to compare)
*Note*: Use a significance level of 0.05 when interpreting your result

**Your answer:**


```{r}

t.test(smoker$weight, nonsmoker$weight)
print(mean(smoker$weight))
print(mean(nonsmoker$weight))

```
**The conclusion of the two-sample test is that we can reject the null hypothesis, AKA, the mean brithweight is not the same between smoking and non-smoking mothers. THe mean birthweight for smoking mothers is lower than that for nonsmoking mothers.**
</div>

The previous hypothesis is to test if the weights of smoker and nonsmoker are
different from each other. It is a two-sided test, since it looks for difference 
in either diretion. If we only want to know if the birthweight of smoker is
*less than* the birthgweight of non-smokers, we can change the null hypothesis to:
the birthweight of smoker is *larger or equal* to the the birthweight of non-smoker.
Remember we mentioned in the lecture the null hypothesis tends to be
something we'd like to reject. This becomes a single-sided test. In other words:

- $H_{0}: \mu_{smoking} \ge \mu_{non-smoking}$; $H_A: \mu_{smoking} \lt \mu_{non-smoking}$ 


To do this test, we can simply use:

```{r}
t.test(smoker$weight, nonsmoker$weight, alternative='less')
```

<div class="question">

**Q6:** Comparing this output with the previous `t.test()`, what do you notice in
terms of *p-value*?

**Your Answer**:  

**The p-value is even lower than our previous test, giving us more confidence to reject the null hypothesis!**

</div>



<div class="question">
 **Q7:** In the dataset, we also know the `gender` of each birth. We'd like to know
if there is significance differences between `male` and `female`. Please
formulate your hypothesis, conduct the hypothesis test and describe your
decision (reject or fail to reject the hypothesis). Use significance level 0.05

**Your answer:**

$H_0: \mu_{male} \le \mu_{female}$; $H_A: \mu_{male} \gt \mu_{female}$ 
```{r}

boys <- nc_clean %>%
  filter(gender=="male")

girls <- nc_clean %>%
  filter(gender=="female")

t.test(boys$weight, girls$weight, alternative='greater')

```

**Based on the p-value of my one-tailed t-test, I can pretty confidently reject
the null hypothesis, with a p-value of ~0.00004587. Therefore, it seems quite likely
that male babies weigh more than female babies on average.**

</div>

## Hypothesis test of population proportions

In the dataset, the column `lowbirthweight` is a categorical variable (true
or false) indicating if the birth is counted as low birth weight. We can
see if the proportions of `lowbirthweight` are different for smoker and
nonsmoker. In this case, we are conducting testing on population proportions.

We haven't covered this in class, but the interpretation is the same as a t-test!

If p < alpha, we reject the null!

We can start by collecting the statistics using the nice function `group_by`
provided in `dlpyr`:

```{r}
nc %>%
  group_by(nc$lowbirthweight, nc$habit) %>%
  summarise(n())
```

Based on the collected statistics, the function to perform the test is `prop.test()`:

```{r}
# for two-sided test
prop.test(c(92,18), c(781+92,108+18))

# for single-sided test
prop.test(c(92,18), c(781+92,108+18), alternative='less')

```


<div class="question">
**Q8:** Based on the output of `prop.test()`, what is the conclusion of the hypothesis test?
*Note*: Use a significance level (alpha) of 0.05

**Your answer:**

**Based on the outcome of the `prop.test`, the test failed to reject the null hypothesis.**

</div>

## Summary: new commands in this lab

* `t.test()`, a function for generic hypothesis testing 
* `prop.test()`, test on proportions
* Review functions (`group_by()`, `filter()`) in the `dlpyr` package




